



## What are R packages, and how do I use them?

An R package is a collection of objects (e.g. functions and data) that are all related to each other, published into one package. 

- R packages on CRAN = books in a bookstore 
- Install package on local computer = buy the book 
- Load package in R session = take the book off the shelf and open it

## Install an R package

Look for "Console" pane.

- Bottom left of your screen

We'll run the following code to install a package:

```{r, eval=FALSE}
install.packages("tidyverse")
```

## Load an R package

To actually use the package, we need to "get it off the bookshelf". 

- To do this, we will use the `library()` function:

```{r, eval=FALSE}
library(tidyverse)
```

Now we are all set! You can utilize any function from ggplot2 in your code. 



Note: If you quit R Studio, it cleans up your books. It effectively puts the book back on the bookshelf. So the next time you open R Studio, *you need to run `library()` again*. 

If you forget to do so, you'll encounter errors like this:

[Error in theme() : could not find function "theme"]{style="color:red;"}

Here, I tried to run the `theme()` function from the `tidyverse` package, but since I hadn't ran `library(tidyverse)` yet, it gives this error. 

## Best practices

- Put all your `library` statements at the top of your qmd file (or your .R, .rmd file). 
  - This keeps things organized, and allows you to quickly have access to all the packages you need whenever you reopen R Studio. 

- **Never put `install.packages()`** in a .qmd, .rmd, or .R file.     
  - Always run this in the console! If you try to render your file, and you have `install.packages()` somewhere in your file, it will give an error. 